Using SMT engine to generate Symbolic Automata\thanks{This work was partially funded by the Associated Team FM4CPS between INRIA and ECNU, Shanghai}

### 2018-01-15 2017-Xudong-SMT/FM-2018-extended.tex
algo

\section{Generation of Open Automata}
\label{section:implementation}

\TODO{Eric: add a header, reorganise the parts of the algorithm}

In this section we describe the algorithm implementing the pNet
semantics, the interaction with the Z3 SMT solver, and we show the
result on our  example.

\TODO{Eric: rewrite}
We have already have the sketch of computing the open automata \cite{???}.
While in our implementation we does not build explicitly a proof tree for every open transition.
Instead, we repeat applying the Tr1 or Tr2 to generate the sub-transitions.
Tr1 is applied on the leaf (pLTS) simply take the pLTS transitions through the given start state and add the predicate in pNet to generate the open transition.
When applying Tr2 we use two methods, combining and matching, to generate the new predicate of the open transitions in a hierarchical manner
as in this case the composition the subnets brings more constraints for synchronizing behaviors. 
Beside the new generated predicate, the predicates from the subnets are also added as a conjunction with those new predicates.

@@

%Algorithm 1 shows the main structure of the algorithm: 
%the computation repeats applying Tr1 and Tr2 to build all possible
%combinations of behaviors of pLTS and Holes, through all possible
%synchronisation vectors. The resulting open transitions are checked
%for satisfiability by the SMT solver.

%We first conduct combining. For each subnet, we enumerate all the
%possible cases of its working status. The result action of a pLTS or a
%pNet is the external action of the node shows the working status of
%it.
%
%For holes, their behavior is unspecified, it doesn't have an exact
%external presentation of action. We only give an action variable to each hole
%and treat it as hole behavior to suppose working status of the
%hole.
%% Here should have a constraint that this variable belongs to the sort of hole.
%In every case, the hole behaviors are unchanged, so we
%only make the combinations of subnets' open transitions. 
%
%Algorithm 1 shows the combining algorithm for enumerating all the possible combinations of open transitions. The algorithm initializes an empty list $L_C$. Dealing with the list $TR$ containing several sets of open transitions from different subnets, the algorithm only chooses one of sets $L_{ot}$ and does combining on the remaining part of $TR$ recursively then get a partial result of combining, $L'_C$. Since there could exist the case that the subnet is not working, a $null$ is added into the $L_{ot}$ to represent it. We get much more combinations after combining $ot$ from $L_{ot}$ with the partial result.

%\begin{algorithm}
%\caption{Combining}
%\begin{algorithmic}[1]
%
%%\Require The list $TR$ of the open transition sets of the subnets. 
%%\Ensure The list of all the possible combinations of open transitions $L_C$.
%\State Initialize an empty combination list $L_C$;
%\State Extract a list of open transitions $L_{ot}$ from $TR$;
%\State Add $null$ to $L_{ot}$;
%\State Get the combinations $L'_C$ of the $TR$;
%\For{each $ot \in L_{ot}$}
%	\For{each $C' \in L'_C$}
%		\State Add $ot$ into $C'$ to get a new tuple $C$;
%		\State Add $C$ into $L_C$;
%	\EndFor
%\EndFor 
%\State \Return $L_C$;
%
%\end{algorithmic}  
%\end{algorithm}

\paragraph{Combining:}
\QIN{
The combining method enumerate all the possible status of the subnets.
The result of the method hence call the combination as it present as all the possible combinations of their open transitions.
Assume that there is a collection of $n$ subnets $L = {s_1, s_2, ..., s_n}$, if we notate $ot$ to be the set of open transitions of the subnet $s$ and $\eta$ means the subnet is not involved. 
Then the combination $\texttt{COMB}$, a set of n-tuples, can be computed as an outer product of sets : 
\[\texttt{COMB} : (\eta\cup ot_1) \times (\eta\cup ot_2)\times ... \times (\eta\cup ot_n) .\]
}
%Algorithm 2 shows the combining algorithm for enumerating all possible
%working status combinations from subnets. Dealing with a list $TR$
%containing  sets of open transitions from different subnets,
%the algorithm choses one of sets $L_{ot}$ and does combining on the
%remaining part of $TR$ recursively. The case when a subnet is not working is
%managed by adding a $null$ into the $L_{ot}$.

\paragraph{Matching:} 
\QIN{$Given a synchronization vector ~ SV_k\in SV = (a'_i)^{i\in I} (b'_j)^{j\in J}\rightarrow v' , G_k , k\in K$, and a tuple
$C_n\in \texttt{COMB} = {(v_i)}^{i\in I} , n\in N$ where $K$ and $N$ are the indices of the set elements.
As the behavior of subnets should match with the synchronization vector to compose the subnets.
The matching method tries matching $SV_k$ and $C_n$ to generate the predicates.
According to the definition of predicate, the hole behaviors $Hole$ and result action $v$ are also involved. 
}

\QIN{
$$Matching(SV_k, C_n, Hole, v) = 
\begin{array}{l}
\forall {(a'_i)}^{i\in I},
{(b'_j)}^{j\in J},v'.
\\
v_i=a'_i\land \, b_j=b'_j \land v=v' \land G_k
\end{array}$$
which is exactly the new part of predicate for one possible result.
}

\paragraph{Filtering:}
\TODO{Eric: 3 levels of filtering, explain our choice}

@@

For example, in the running example, the first possible behavior of
the hole, it doesn't have a name, only have the prefix ``hb''. The
hole is the second subnets of the root node so its ``tree index'' is
``12''. The fresh variable name is ``$\OTvar{:hb:12:1}$''. 
\subsection{Variable managment}
The variables in each synchronisation vector are considered local: for a given pNet expression, we must have fresh local variables for each occurrence of a vector (= each time we instantiate rule Tr2). Similarly the state variables of each copy of a given pLTS in the system, must be distinct, and those created for each application of Tr2 have to be fresh and all distinct. This will be implemented within the open-automaton generation algorithm, e.g. using name generation using a global counter as a suffix.
\subsection{Filtering the behaviour combinations}
\QIN{There is already able to filter some wrong open transitions during the matching.
We can ensure that the synchronization vector is not matched with the subnets if there is only one side not involved.
Then we can remove the transitions from the result of matching according to the condition $\gamma$ where
$\gamma = \bigcup_{i\in I}(v_i\in\eta\land a'_i\in\eta)\lor(v_i\in\eta\land a'_i\in\eta) .$
}
\subsection{Translation to SMTlib input language}
In order to submit satisfiability problems to Z3 (for the predicates
in open transitions), we need to generate SMTlib programs, from the
pNet Algebra presentation and predicates.
More precisely, we need to translate to SMTlib:
\begin{itemize}
  \item the presentation of the action algebra
    (sorts and operators) that is defined for a given language (process calculus, or high level programming language),
    \item for a given pNet, the set of local constants (actions or auxiliary data) that are used in the pLTSs,
    \item for each open transition: the declaration of variables, and
      the predicate (including action expressions).
\end{itemize}

>>>>

\section{Generation of Open Automata}
\label{section:implementation}
In this section we describe the algorithm implementing the pNet
semantics, the interaction with the Z3 SMT solver, and we show the
result on our  example.

Our algorithm starts with an open pNet, and builds its set of open
transitions (OTs). Its main loop is a classical residual algorithm: starting
from the initial global state, it pick a state in an unexplored set, and
computes all possible OTs, adding their target states in the
unexplored set, until this set is empty. As the potential set of
global states is finite (see \cite{henrio:Forte2016}), this terminates.

The inside loop (\emph{MakeTransitions} method) applies recursively
the semantic rules following the structure of the pNet. 
For pLTS at leaves, rule Tr1 simply take the pLTS transitions of the
corresponding local state and build the predicate to generate the
open transition. 
On a pNet node to apply rule Tr2 we use two methods, combining and
matching, to generate the new predicate of the open transitions in a
hierarchical manner.

Last, at the root of the pNet, the predicate of each OT is translated
into SMTlib assertions, and checked for satifiability. Unsatisfiable
OTs are pruned, and the final open automaton is computed as the set of
global states reachable by satisfiable OTs.

@@

%
%For holes, their behavior is unspecified, it doesn't have an exact
%external presentation of action. We only give an action variable to each hole
%and treat it as hole behavior to suppose working status of the
%hole.
%% Here should have a constraint that this variable belongs to the sort of hole.
%In every case, the hole behaviors are unchanged, so we
%only make the combinations of subnets' open transitions. 
%

\def\inactive{\{-\}}
\paragraph{Combining:}
The combining method enumerate all the possible behaviours of the
subnets, as all the possible combinations of their open transitions.
Assume that there is a collection of $n$ subnets $L = {s_1, s_2, ...,
  s_n}$, if we notate $ot_i$ to be the set of open transitions of the
subnet $s_i$ and ``$-$'' means the subnet is not involved.  
Then the combination $\texttt{COMB}$, a set of n-tuples, is the
cartesian product of sets :  
\[\texttt{COMB} : (\inactive\cup ot_1) \times (\inactive\cup ot_2)\times ... \times (\inactive\cup ot_n) .\]

%Algorithm 2 shows the combining algorithm for enumerating all possible
%working status combinations from subnets. Dealing with a list $TR$
%containing  sets of open transitions from different subnets,
%the algorithm choses one of sets $L_{ot}$ and does combining on the
%remaining part of $TR$ recursively. The case when a subnet is not working is
%managed by adding a $null$ into the $L_{ot}$.

\paragraph{Matching:}
The next step, for each synchronisation vector, and each possible
combination of behaviour of the subnets as generated by the \emph
{Combining} method, consists in building the combined predicate,
and the corresponding open transition.
From a synchronization vector
$sv_k\in SV = (a'_i)^{i\in I} (b'_j)^{j\in J}\rightarrow v'$
and its guard $G_k$, a tuple of open transitions
$C_n\in \texttt{COMB} = {(v_i)}^{i\in I}$, the hole behaviors $Hole$,
and a fresh variable representing the result action $v$, we build the
predicate:

$$Matching(SV_k, C_n, Hole, v) = 
\begin{array}{l}
\forall {(a'_i)}^{i\in I},
{(b'_j)}^{j\in J},v'.
\\
v_i=a'_i\land \, b_j=b'_j \land v=v' \land G_k
\end{array}$$

\paragraph{Filtering:}
There may be several reasons why such a matching would fail:
\begin{itemize}
  \item if some subnet is marked as inactive in the vector, and the
    chosen combinaison has an active behaviour at this position,
  \item if some subnet action expression in the vector does not match (by
    pattern-matching) the corresponding behaviour $C_n$,
  \item or if the whole set of active subnet actions in the vector
    cannot be matched (by unification) with the set $\{C_n\}$.
\end {itemize}

Each of these possible failures is more precise than the previous
ones. But even when unification succeeds, it is still possible that
the resulting predicate would be unsatisfiable, because of some
incompatibility involving the guards. In our algorithm, we choose to
apply only the simplest filter inside the \emph{Matching} method. All
the rest will be checked later, using the satisfiability ckeck in the
SMT engine.

@@

For example, in the running example, the first possible behavior of
the hole, it doesn't have a name, only have the prefix ``hb''. The
hole is the second subnets of the root node so its ``tree index'' is
``12''. The fresh variable name is ``$\OTvar{:hb:12:1}$''. 
\subsection{Variable managment}
The variables in each synchronisation vector are considered local: for a given pNet expression, we must have fresh local variables for each occurrence of a vector (= each time we instantiate rule Tr2). Similarly the state variables of each copy of a given pLTS in the system, must be distinct, and those created for each application of Tr2 have to be fresh and all distinct. This will be implemented within the open-automaton generation algorithm, e.g. using name generation using a global counter as a suffix.

\subsection{Translation to SMTlib input language}
In order to submit satisfiability problems to Z3 (for the predicates
in open transitions), we need to generate SMTlib programs, from the
pNet Algebra presentation and predicates.
More precisely, we need to translate to SMTlib:
\begin{itemize}
  \item the presentation of the action algebra
    (sorts and operators) that is defined for a given language (process calculus, or high level programming language),
    \item for a given pNet, the set of local constants (actions or auxiliary data) that are used in the pLTSs,
    \item for each open transition: the declaration of variables, and
      the predicate (including action expressions).
\end{itemize}
